Smart Representations: {R}ationality and Evolution in a Richer Environment

### 2016-06-24 paper/rev1_paperBlinded.tex
revision version by Paolo

@@

\begin{abstract}
\normalsize
  Standard applications of evolutionary game theory look at a single, fixed game and focus on
  the evolution of behavior for that game alone. Instead, this paper uses tools from
  evolutionary game theory to study the evolutionary competition between \emph{choice
    mechanisms} in a rich and variable environment. A choice mechanism is a way of representing
  a decision situation, paired with a method for choosing an act based on this 
  subjective representation. We demonstrate the usefulness of this approach by a case study
  that compares a number of classic proposals for decision making. Our
  main result is that regret-based preference representations can be evolutionarily selected
  for, despite being non-veridical.
\end{abstract}

@@

In full generality, a choice mechanism associates decision situations with action choices. A crucial
part of a choice mechanism is the \emph{subjective representation} of the decision situation,
in particular the manner of forming preferences and beliefs about a possibly uncertain world (see
figure~\ref{fig:ChoiceMechanism} below). This paper asks: which preference and belief
representations are ecologically valuable and lead to high fitness?
The evolution of
preferences has been subject of recent interest in theoretical economics
\citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. A contribution of this paper is
to provide a case that questions of preference evolution should take
variability in uncertainty representation into account as well. As a point in case, we demonstrate
that if agents have imprecise probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96},
preference representations in terms of regret can be selected for, even though they may deviate
from the true, objective fitness that evolutionary selection operates on.

@@

Section \ref{sec:rati--subj} reviews different views of rational choice and relates the
approach we take here to the existing literature. Section \ref{sec:basic-notions} introduces the
relevant decision- and game-theoretic notions, and
%spells out four choice mechanisms on whose
%evolutionary competition we zoom in 
Section~\ref{sec:basic-model-1} presents the
main results of the paper. Section \ref{sec:extensions} extends the evolutionary analysis by
introducing more choice mechanisms and alternative models of the environment.

\section{Rationality, Representations and Meta-games}
\label{sec:rati--subj}

@@

\begin{definition}[Rational choice from maximization of expected utility]
  \label{def:rationality}
  Let $S$ be a (finite) set of states and $A$ a (finite) set of actions. Given a
  probability measure $\mu$ over $S$ and a payoff function
  $u \mycolon S \times A \rightarrow \mathbb{R} $, an action $a^* \in A$ is rational
  if it maximizes the \emph{(subjective) expected utility}
  $\EU(a, \mu) := \sum_{s \in S} \mu(s) \ u(s, a)$.
\end{definition}

\noindent Expected utility is subjective in the sense that it is a function of subjective
beliefs $\mu$ and subjective preferences $u$ of the decision maker (DM). To wit, a choice can be
rational, i.e., the best choice from the DM's point of view, even if based on peculiar beliefs
and/or aberrant preferences. % (Indeed, while
% \citeauthor{Neumannvon-NeumannMorgenstern1944:Theory-of-Games} assumed that probabilities were
% objectively given, it was \citet{Savage1954:The-Foundations} who famously extended von Neumann
% and Morgenstern's axiomatization to include a subjective notion of probabilistic beliefs on top
% of the subjective notion of preference that von Neumann and Morgenstern started out with.)

@@

An alternative view on rationality of choice is \emph{adaptationism}
\citep[e.g.,][]{Anderson1991:Is-human-cognit,ChaterOaksford2000:The-Rational-An,HagenChater2012:Decision-Making}. Adaptationism
aims to explain rational behavior by appeal to evolutionary considerations: DMs have acquired
choice mechanisms that have proved to be adaptive with respect to the variable environment
where they have evolved. A choice mechanism can be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.,][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to closely relate to the literature on evolution of preferences and to the philosophical debate
about the nature of rational choice in general, we here suggest to think of a choice mechanism
as a map from choice situations to action choices which includes an explicit level of
subjective representation of the situation, that we call \emph{subjective representation
  scheme} (see figure~\ref{fig:ChoiceMechanism}). Specifically, a subjective representation
scheme is a general way of forming preferences and beliefs about the choice situation, and we
are most interested in the question which subjective representations are better than others
from an evolutionary point of view.

@@

The fitness of a subjective representation scheme is measured in terms of the fitness of a
choice mechanism that contains it, all else equal. The fitness of a choice mechanism $c$ is the
expected payoff of $c$ over arbitrary decision situations $D$:
\begin{align}
  \label{eq:FittnessChoiceMechSolitary}
  F(c) = \int P(D) \  \pi_D(c) \text{ d} D \,,
\end{align}
where $P(D)$ is the probability (density) of $D$ occurring and $\pi_D(c)$ is the evolutionary payoff of
the act selected by $c$ in $D$. The probability measure $P(D)$ encodes the statistical
properties of the environment.

The same logic also applies to game contexts and social decision making. A game $G$ is also a
choice situation, albeit a more complex one. A choice mechanism $c$ maps $G$ onto an action
choice in $G$. The fitness of $c$ is then:\footnote{We assume for simplicity of exposition that
  all games with positive occurrence probability are symmetric two-player games.}
\begin{align}
  \label{eq:FittnessChoiceMechGame}
  F(c) = \int \int P(G) \  P(c') \  \pi_G(c,c') \text{ d} c' \text{ d} G  \,,
\end{align}
where $P(G)$ is the probability (density) of game $G$, $P(c')$ is the probability of
encountering a co-player who uses choice mechanism $c'$ and $\pi_G(c,c')$ is the evolutionary payoff for
a $c$-player in $G$ against the $c'$-player, given by the actions that $c$
and $c'$ select in $G$. For the most part, we will focus on the more complex case of game situations. Results about solitary decision making follow from parallel arguments
(see Section~\ref{sec:extensions}).

\emph{Meta-games} are abstract models for the evolutionary competition between choice
mechanisms in interactive decision making contexts. The expected payoff for a $c$-player in a random
encounter with a $c'$-player is the average of payoffs obtained in a random symmetric game $G$ weighted
by $G$'s occurrence probability:
\begin{align}
  \label{eq:FittnessChoiceMechGamePairwise}
  F(c, c') = \int P(G) \  \pi_G(c,c') \text{ d} G \,.
\end{align}

\begin{definition}[Meta-game]
  \label{def:MetaGame}
  Let $CM$ be a set of choice mechanisms and let $P(G)$ give the occurrence probability of symmetric game
  $G$ in the environment. The meta-game that captures selection of choice mechanisms in this
  environment is the one-population evolutionary game with strategies $CM$ and payoff function
  $\pi \mycolon CM \times CM \rightarrow \mathbb{R}$ such that
  $\pi(c,c') = F(c,c')$.
  % = \int P(G) \  \pi_G(c,c') \text{ d} G$.
\end{definition}

Standard notions of evolutionary game theory can be applied to meta-games as well. A choice
mechanism $c$ is a strict Nash equilibrium if $F(c,c) > F(c',c)$ for all $c'$; it is
evolutionarily stable if for all $c'$ either (i) $F(c,c) > F(c',c)$ or (ii) $F(c,c) = F(c',c)$
and $F(c,c') > F(c',c')$; it is neutrally stable if for all $c'$ either (i) $F(c,c) > F(c',c)$
or (ii) $F(c,c) = F(c',c)$ and $F(c,c') \ge F(c',c')$
\citep{Maynard-Smith1982:Evolution-and-t}. Similarly, evolutionary dynamics can be applied to
meta-games. We will later turn towards replicator dynamics
\citep{TaylorJonker1978:Evolutionary-St} and replicator mutator dynamics
\citep[e.g.][]{Nowak2006:Evolutionary-Dy}.
%\todo{footnote with caveat about agent-level processes}

\section{Classic Choice Rules and Choice Mechanisms}
\label{sec:basic-notions}

@@

\subsection{Classic Rules for Decision Making}
\label{sec:choice-rules}

@@

A strategic-form game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a finite set of
players, $A_i$ a finite set of acts and $\pi_i: \bigtimes_{j \in N} A_j \rightarrow \mathbb{R}$ a payoff function for each player $i \in N$.
Part of this paper will focus on symmetric games with two players. A two-player game is
symmetric if players have the same action set, $A_1 = A_2$, and payoffs are symmetric:
$\pi_1(a_k, a_j) = \pi_2(a_j, a_k)$ for all $a_j,a_k \in A_1$. An example is given in
figure~\ref{coordgame1}. The numbers give the payoff for the player choosing a row act against
a co-player choosing a column act. Since the game is symmetric, it suffices to give just these
numbers.

Rational choice from maximization of expected utility, as given in
Definition~\ref{def:rationality}, applies to games as follows. To say which acts are rational,
it requires specifying a precise probabilistic belief of DM. Let $A_{-i} := \bigtimes_{j \neq i} A_j$ be
the set of all \emph{profiles} of acts for all players except $i$. If player $i$ has a belief
$\mu_i \in \Delta(A_{-i})$ about the behavior of his co-players, he chooses rationally only if
his choice $a^*_i$ maximizes expected utility
$\EU(a_i, \mu) := \sum_{\act_{-i} \in \Acts_{-i}} \mu_i(\act_{-i}) \myts \pi_i(a_i, a_{-i})$. For
the example in figure~\ref{coordgame1}, an expected utility maximizer would choose act $I$ only if he
believes that his co-player is going to choose act $I$ with a probability of at least
$\nicefrac{2}{3}$.

Perhaps the most well-known classic rule that does not require precise probabilistic
beliefs is the \emph{maxmin rule}. It prescribes to choose act $\act^*$ only if it
maximizes payoffs for all worst case scenarios:
$\act^* \in \argmax_{a_i \in \Acts_i} \min_{a_{-i} \in \Acts_{-i}} \pi_i(a_{i},
a_{-i})$.
For the example in figure~\ref{coordgame1}, the minimum of both acts is 0: players choosing by maxmin
rule would be indifferent between acts $I$ and $II$.

\emph{Laplace rule} selects $\act^*$ if
$\act^* \in \argmax_{a_{i} \in \Acts_{i}} \sum_{a_{-i}} \pi(a_{i}, a_{-i})$.
Laplace rule is equivalent to expected utility maximization under a \emph{flat belief} that
assigns equal probability to all of the co-players' choices. In the coordination game of
figure~\ref{coordgame1}, Laplace rule selects act $II$.

The final classic rule that is relevant for our purposes is \emph{regret
  minimization}. The notion of regret in decision theory dates back at least to the work by
\citet{Savage1951:The-theory-of-s}, and has later been developed by \citet{bell82}, \citet{fish82} and \citet{loosug82} independently. Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can give
solutions to game theoretical-puzzles (like the Traveller's dilemma and the Centipede game) in a way that is closer to
everyday intuition and empirical data. In this paper the notion of regret is defined as in
\citet{HalpernPass2012:Iterated-Regret}:

\begin{definition}[Regret, Regret Minimization] \label{defn:regret} Given a game
  $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the \emph{pairwise regret} of player $i$'s
  act $a_i$ against profile $a_{-i}$ is
  $\text{preg}(a_i,a_{-i}):= \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $, where
  $a_i^\$ \in \argmax_{a_i \in A_i} \pi_i(a_i, a_{-i})$ denotes $i$'s best reply to the
  co-players' actions $a_{-i}$. The \emph{regret} of $a_i$ is then
  $\text{reg}(a_i):= \text{max}_{a_{-i}\in A_{-i}} \text{preg}(a_i,a_{-i}) $. An
  action $a^{*}_i $ is regret minimizing if
  $a^{*}_i \in \text{argmin}_{a_i} \text{max}_{a_{-i}} \text{preg}(a_i,a_{-i})$.
\end{definition}

\noindent To illustrate, figure~\ref{coordgame1reg} shows the pairwise regrets and the regrets for the game of figure~\ref{coordgame1}. 
We have
$\text{reg}(I)=2$, because $\text{preg}(I,II)=2$, whereas $ \text{preg}(I,I)=0 $. With the same reasoning, $\text{reg}(II)=1$. A regret minimizer would therefore choose act $II$.

\begin{figure}

  \begin{subfigure}[b]{0.3\textwidth}
    \centering
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & 1 & 0 \\
      II & 0 & 2\\
      \bottomrule
    \end{tabular}
    \caption{Coordination game $G^C$}
    \label{coordgame1}
  \end{subfigure}
  \hspace{1cm}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \begin{tabular}{ccc|c}
      \toprule
      & $\text{preg}( \cdot, I)$ & $\text{preg}(\cdot, II)$ & $\text{reg}(\cdot)$ \\
      \midrule
      I  & 0 & 2 & 2 \\ 
      II & 1 & 0 & 1\\
      \bottomrule
    \end{tabular}
    \caption{Pairwise regrets and regrets}
    \label{coordgame1reg}
  \end{subfigure}
  \caption{A coordination game (left) and regrets associated with the row player's acts (right).}
    \label{coordgame1mainFig}
\end{figure}

% \begin{example}[The Traveller's Dilemma \citep{HalpernPass2012:Iterated-Regret}]
% \label{example:TravelersDilemma}
% Let us now consider the Traveller's Dilemma case. The story behind the game goes as
% follows. There are two travellers, Ann and Bob, who lost their luggages. Ann and Bob had
% exactly the same luggage and they had insured the luggages with the same insurance company. The
% insurance policy is that the two travellers have to separately claim a certain amount between
% 2\$ and 100\$ for the reimbursement, and if they both claim the same amount then they will be
% reimbursed by that amount. Otherwise, if one of them claims more than the other, then the one
% who claimed the higher amount will get the lower amount minus 2\$, while the one who claimed
% the lower amount will get what (s)he claimed plus 2\$. The game is symmetric, and the payoff
% function for both players is:
% \begin{align*}
%   \pi_i(a_i,a_{-i}) = \begin{cases} a_i \text{ \hspace{1cm} if } a_i = a_{-i} \\ a_i + 2 \text{
%       \hspace{.4cm} if } a_i < a_{-i} \\ a_{-i} -2 \text{ \hspace{.25cm} if } a_{-i} <
%     a_i \end{cases}
% \end{align*}
% By a simple computation we have:  $\forall a_i \in \lbrace 96\$, ..., 100\$ \rbrace \text{, }
% \text{reg}(a_i)=3 $ and $\forall a_i \in \lbrace 2\$, ..., 95\$ \rbrace \text{, }
% \text{reg}(a_i)>3$. If we now eliminate all the actions that do not minimize regret in the
% first place and iterate the regret minimization only on the action set $\lbrace 96\$, ...,
% 100\$ \rbrace$, we get that 97\$ is the action that minimize regret. Indeed, $\text{reg}(97)=2$
% and $\forall a_i \in \lbrace 96\$, 98\$, 99\$, 100\$ \rbrace \text{, } \text{reg}(a_i)=3
% $. Hence, one step of regret minimizing reasoning eliminates all the actions smaller than 96\$,
% and if we iterate the process once more the unique outcome is 97\$. Finally, it is worth
% noticing that the only rationalizable equilibrium of the game corresponds to the outcome
% $(2,2)$, that is consequently the only Nash equilibrium, the only perfect equilibrium, and the
% only sequential equilibrium of the game.

% \begin{itemize}
% \item \textcolor{red}{add here:}
%   \begin{itemize}  
%   \item outcome of rationalizability (max-EU)
%   \item outcome of maxmin rule
%   \item outcome of Laplace rule
%   \end{itemize}

% \end{itemize}

%  $ \medsquare $

% \end{example}

Regret minimization is intimately related to the maxmin rule. Let the \emph{negative regret
  transformation} of a strategic game $G$ be the game $G^r$ derived from $G$ by replacing the payoff
functions $\pi_i$ of $G$ with the negative pairwise regrets $-\text{preg}(a_i,a_{-i})$. For
example, the negative regret transformation of the game in figure~\ref{coordgame1} would have the
following payoff matrix, i.e., the negatives of the pairwise regrets given in
figure~\ref{coordgame1reg}:

\begin{center}
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & 0 & -2 \\
      II & -1 & 0\\
      \bottomrule
    \end{tabular}
\end{center}

\noindent Then, for any game $G$, acts selected by maxmin rule for the negative regret transformation of $G$ are
always exactly the acts selected by regret minimization for the original game $G$.
This means that we can think of the maxmin rule and regret minimization as essentially the same
action selection function on different subjective representations of preferences: while maxmin considers actual
objective payoffs, regret minimization considers non-veridical, regret-based subjective
utilities. As the example of figure~\ref{coordgame1} shows, different subjective preferences
can give rise to different choices, all else equal. While maxmin is indifferent
between $I$ and $II$, regret minimization uniquely selects $II$.

In sum, classic proposals for rational decision making are normative prescriptions of action choices
for DM. Such prescriptions may rest on implicit assumptions about what DM
believes (e.g., when interpreting Laplace rule as expected utility maximization under flat beliefs). They may
also make implicit assumptions about what DM prefers (e.g., when considering regret-based preferences). This is rather messy. To disentangle, we rephrase all the above in terms of choice mechanisms.

\subsection{Choice Mechanisms}
\label{sec:choice-mechanisms}

@@

The subjective representation scheme $\gamma_i$ of agent $i$ is a function that
takes a game $G = \tuple{ N, (A_i , \pi_i)_{i \in N} }$ and maps it onto
$\gamma_i(G) = (\theta, e)$ where
$\theta \mycolon \bigtimes_{j \in N} A_j \rightarrow \mathbb{R}$ is a subjective utility function (the \emph{preference type} of the player), and $e$ is some subjective representation of a player's uncertainty about the co-players'
behavior (the \emph{epistemic} or \emph{belief type} of the player). As already anticipated, we mostly consider two preference types here: the \emph{objective type} maps each game
onto its own, true evolutionary payoffs (i.e., $ \theta = \pi_i $); the \emph{regret type} maps each game onto its negative regret
transformation (i.e., $ \theta = -\text{preg} $). We also consider two belief types: the
\emph{precise type} represents uncertainty about the co-players in a single probability measure
$\mu \in \Delta(\Acts_{-i})$; the \emph{imprecise type}
represents uncertainty as a set of probability measures $\Gamma \subseteq \Delta(\Acts_{-i})$.
It is convenient to conceptualize precise beliefs as a special case of imprecise beliefs by
looking at $\mu$ as a stand-in for the singleton set $\set{\mu}$.
The action selection function $\kappa_i(\gamma_i(G)) \subseteq A_i$ that we consider here is \emph{maximinimization of expected utility} (see Appendix~\ref{sec:impr-prob-beli} for motivations).

\begin{definition}[Maximinimization of expected utility]
  \label{def:maxminSEU}
  Fix a game $G$. Let agent $i$'s subjective representation scheme
  $\gamma_i(G) = (\theta, e)$ be such that $\theta$ is $i$'s subjective utility function and $e \subseteq \Delta(\Acts_{-i})$ is a (possibly singleton) set of probability measures. Agent $i$'s expected utility for act $a_{i}$ and belief
  $\mu \in e$ is
  $\EU(a_{i}, \mu) = \sum_{a_{-i} \in \Acts_{-i}} \mu(a_{-i}) \theta(a_{i},
  a_{-i})$.
  Act $a_{i}^*$ is selected by maximinimization of expected utility iff
  $a_{i}^* \in \arg \max_{a_{i} \in \Acts_{i}} \min_{\mu \in e} \EU(a_{i}, \mu)$.
\end{definition}

\iffalse
\begin{fact} \label{fact:singleton probability set}
Maximinimization under a precise (singleton set) belief is just standard
maximization of expected utility (Definition~\ref{def:rationality}).
\end{fact}
\fi

\begin{fact} \label{fact:maxEU-minReg} 
%Under an identical precise (singleton set) belief, the
%  acts selected by maximization of expected evolutionary payoff in game $G$ are exactly the acts selected
%  by the maximization of expected utility for the negative regret transformation of $G$
Under an identical precise (singleton set) belief, maximization of expected evolutionary payoff and minimization of expected regret coincide. %\citep[e.g.,][]{HalpernPass2012:Iterated-Regret}.
\end{fact}

@@

Under maximal uncertainty, all four choice mechanisms considered here are behaviorally
equivalent on $2 \times 2$ symmetric games, except for objective types with
imprecise beliefs:

\begin{fact} \label{fact:equivalence2x2} In the class of $2 \times 2$ symmetric games, the acts
  selected by Laplace rule are exactly the acts selected by regret minimization.
\end{fact}

\section{The Model}
\label{sec:basic-model-1}

>>>>

\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{calc}

@@

\begin{abstract}
\normalsize
  Standard applications of evolutionary game theory look at a single, fixed game and focus on
  the evolution of behavior for that game alone. Instead, this paper uses tools from
  evolutionary game theory to study the evolutionary competition between \emph{choice
    mechanisms} in a rich and variable environment. A choice mechanism is a way of representing
  a decision situation, paired with a method for choosing an act based on this 
  subjective representation. We demonstrate the usefulness of this approach by a case study
  that compares a number of classic proposals for decision making. The
  main result is that regret-based preference representations can be evolutionarily selected
  for, despite being non-veridical.
\end{abstract}

@@

In full generality, a choice mechanism associates decision situations with action choices. A crucial
part of a choice mechanism is the \emph{subjective representation} of the decision situation,
in particular the manner of forming preferences and beliefs about a possibly uncertain world (see
figure~\ref{fig:ChoiceMechanism} below). This paper asks: which preference and belief
representations are ecologically valuable and lead to high fitness?
The evolution of
preferences has been subject of recent interest in theoretical economics
\citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. Here, we argue that questions of preference evolution should take
variability in uncertainty representation into account as well. In relation to this, 
the main contributions of this work are essentially three. First, we show that the way agents form beliefs, and the type of uncertainty that they entertain in particular, is a key aspect for the evolutionary success. 
Second, we demonstrate
that if agents have imprecise probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96}, faithful and objective representation in terms of true, real evolutionary fitness can be outperformed by
subjective (e.g., regret-based) preference representations that deviate
from the true fitness that natural selection operates on.
Third, although a few examples along similar lines have recently been introduced (REFERENCES), our approach based on a meta-game model is very uncustomary in the literature, and contributes to widen the methods of evolutionary game theory beyond the traditional single-game framework.

@@

The paper is organized as follows. Section \ref{sec:rati--subj} reviews different views of rational choice and relates the
approach we take here to the existing literature. Sections \ref{sec:subj-ut,bel,dec-rules} and \ref{sec:subj-rep,ev-fit} introduce the
relevant decision-theoretic and game-theoretic notions, while Sections \ref{sec:metagame} and \ref{sec:model} present the model.
%spells out four choice mechanisms on whose
%evolutionary competition we zoom in 
The main results of the paper are in Section~\ref{sec:results}, and Section \ref{sec:extensions} extends the evolutionary analysis by
introducing more choice mechanisms and alternative models of the environment. Finally, Section \ref{sec:conclusion} concludes.

\section{Rationality and Representations}
\label{sec:rati--subj}

@@

\noindent Expected utility is subjective in the sense that it is a function of subjective
beliefs and subjective preferences of the decision maker (DM). To wit, a choice can be
rational, i.e., the best choice from the DM's point of view, even if based on peculiar beliefs
and/or aberrant preferences. % (Indeed, while
% \citeauthor{Neumannvon-NeumannMorgenstern1944:Theory-of-Games} assumed that probabilities were
% objectively given, it was \citet{Savage1954:The-Foundations} who famously extended von Neumann
% and Morgenstern's axiomatization to include a subjective notion of probabilistic beliefs on top
% of the subjective notion of preference that von Neumann and Morgenstern started out with.)

@@

An alternative view on rationality of choice is \emph{adaptationism}
\citep[e.g.,][]{Anderson1991:Is-human-cognit,ChaterOaksford2000:The-Rational-An,HagenChater2012:Decision-Making}. Adaptationism
aims to explain rational behavior by appeal to evolutionary considerations: DMs have acquired
choice mechanisms that have proved to be adaptive with respect to the variable environment
where they have evolved. A choice mechanism can be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.,][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to closely relate to the literature on evolution of preferences and to the philosophical debate
about the nature of rational choice, we here suggest to think of a choice mechanism
as a map from choice situations to action choices which includes an explicit level of
\emph{subjective representation} of the situation. Specifically, a subjective representation is a general way of forming preferences and beliefs about the choice situation, and we
are most interested in the question which subjective representations, and which choice mechanisms in general, are better than others
from an evolutionary point of view.

\iffalse

@@

@@

@@

@@

@@

\fi

\section{Subjective Utilities, Beliefs and Decision Rules} \label{sec:subj-ut,bel,dec-rules}

We view a choice mechanism as the combination of three different things: a \textit{subjective utility} (or preference), a \textit{subjective belief} and a \textit{decision rule}. In general, the agent's action choice will depend both on the agent's utility at different possible outcomes of the choice situation and on the agent's beliefs about the realization of these outcomes. The decision rule then combines the agent's subjective utility and belief, and dictates how the agent should act in case of a choice situation whose consequences are uncertain. Therefore, a decision rule is a function that associates an action choice with the agent's utility and beliefs:
$$ \text{Decision Rule: Utility }\times\text{ Beliefs } \rightarrow \text{ Actions.}  $$
The utility of an agent can be formally expressed by a function $u:W \times A \rightarrow \mathbb{R} $, where $A$ stands for a (finite) set of actions available to the agent and $W$ is a (finite) set of possible states of the world. There are many different ways to describe beliefs, but here we assume that the agent's beliefs are represented in terms of a (possibly singleton) set of probability functions $\Gamma \subseteq \Delta(W) $ over the possible states of the world. In line with this definition, we can consider different decision rules, such as:
\begin{enumerate}

\item \textbf{Maxmin}: $$ a^*(u,\Gamma)= \argmax_{a \in A} \min_{\mu \in \Gamma} \sum_{w\in W} u(w,a)  \mu(w).$$

\item \textbf{Maximax}: $$ a^*(u,\Gamma)= \argmax_{a \in A} \max_{\mu \in \Gamma} \sum_{w\in W} u(w,a)  \mu(w).$$

\item \textbf{Laplace rule}: $$ a^*(u,\Gamma)= \argmax_{a \in A} \sum_{w\in W} \frac{1}{|W|}  u(w,a) .$$

\item \textbf{Expected utility maximization} (for $\Gamma$ singleton): $$ a^*(u,\Gamma)= \argmax_{a \in A} \sum_{w\in W} u(w,a)  \mu(w).$$

\end{enumerate}

It is worth noticing that both maxmin and maximax boil down to expected utility maximization when the set  $\Gamma$ is a singleton, and in turn expected utility maximization reduces to Laplace rule when the belief $\mu$ is a uniform probability over the states.

\section{Subjective Representations and Evolutionary Fitness} \label{sec:subj-rep,ev-fit}

For a choice mechanism to prescribe an action, the decision rule  needs to be given a specific subjective utility and belief as input, as described in the previous section. We then call \textit{subjective representation} of the decision problem the pair $(u,\Gamma)$ of a subjective utility and a subjective belief.
In other words, a subjective representation is a general way of forming preferences and beliefs about the choice situation.

The literature on evolution of preferences (REFERENCES) mainly focuses on the evolutionary selection of subjective utilities uniquely. Instead, it will be clear from our model that different subjective beliefs may have a fundamental role in determining the evolutionary success of subjective utilities, and therefore the interplay between utilities and beliefs needs to be taken into account explicitly.
So, here we
are most interested in the question which subjective representations are better than others
from an evolutionary point of view. In order to investigate this issue we are going to use the machinery of evolutionary game theory.

A game is an interactive decision situation. 
For a given game $G=\langle N, (A_i,\pi^G_i)_{i \in N} \rangle$, let us denote the evolutionary payoff, or \textit{fitness}, of player $i$ by the function $\pi^G_i: \Pi_{i \in N} A_i \rightarrow \mathbb{R}$, where $A_i$ is player $i$'s (finite) set of actions. For simplicity of exposition we assume that
  all games that are played are symmetric two-player games, that is, games such that: $N:=\{1,2\}$, $A_1 = A_2 $, and $\pi^G_1(a_1, a'_2)= \pi^G_2(a'_1, a_2)=:\pi^G(a,a')$.\footnote{Since payoff functions are symmetric, we will simply write $\pi^G(a, a')$ for $\pi^G_1(a_1, a'_2)$ and $A:=A_1=A_2$, as usual. However, notice that all the definitions and results can be extended to more general cases.} Then, the evolutionary fitness of a subjective representation $(u,\Gamma)$ is measured in terms of the fitness of a
choice mechanism that contains it, all else equal. 
Formally, the fitness of a choice mechanism $c$ against choice mechanism $c'$ in a symmetric two-player game $G=\langle \{1,2\}, A, \pi^G \rangle$ is the
expected payoff of $c$ against $c'$ in $G$:

$$
 F_G(c,c')= \pi^G(a_c^*(u^G,\Gamma),a_{c'}^*(u^G,\Gamma)), \footnote{Whenever a choice mechanism would not select a unique
  action, we assume that the player chooses one of the equally optimal actions at random. I.e., $F_G(c,c')= \sum_{a \in a_c^*(u^G,\Gamma)} \sum_{a' \in a_{c'}^*(u^G,\Gamma)} \frac{1}{|a_c^*(u^G,\Gamma)|} \frac{1}{|a_{c'}^*(u^G,\Gamma)|} \pi^G(a,a')$.}
$$
where $a^*_c$ is the decision rule of choice mechanism $c$. Given the game theoretic setting, the subjective utility $u$ is now a function $u^G:A \times A \rightarrow \mathbb{R}$, and the subjective belief $\Gamma$ is a set of probability functions over the co-player's actions, $ \Gamma \subseteq \Delta(A) $.

However, studying the fitness and the evolutionary selection of different choice mechanisms by the use of a fixed, single game may look like a rather restrictive perspective and possibly biased by the choice of the particular game. To avoid these limitations, we extend the standard single-game framework of evolutionary game theory by considering a \textit{class} of possible games that can be played within a population. Although not completely novel (seeREFERENCES), this approach has started being investigated only very recently, and we maintain that it can shed new light on the evolution of behavior, and it can open to much broader perspectives on evolutionary game theory in general.

\section{The Meta-Game} \label{sec:metagame}

The ground for the evolutionary competition of choice mechanisms will consist of a class $\mathcal{G}$ of possible two-player symmetric games, together with a probability measure $P_{\mathcal{G}}(G)$ for the occurrence probability of game $G \in \mathcal{G}$. In more biological terms, the probability measure $P_{\mathcal{G}}$ encodes the statistical
properties of the environment. We then call \textit{meta-game} the tuple $ MG=\langle CM, \mathcal{G}, P_{\mathcal{G}},F \rangle$, where $CM$ is a set of choice mechanisms, $\mathcal{G}$ is a class of possible games, $P_{\mathcal{G}}(G)$ is the probability of game $G$ to occur, and $F:CM \times CM \rightarrow \mathbb{R}$ is the fitness function, defined as:
\begin{align}
  \label{eq:FittnessChoiceMechGamePairwise}
  F(c, c') = \int P_{\mathcal{G}}(G) \  F_G(c,c') \text{ d} G \,.
\end{align}
Hence, $F(c,c')$ determines the evolutionary payoff of choice mechanism $c$ against $c'$ in the meta-game. The set $CM$ is though of as the set of choice mechanisms present within a given population that is playing the games from the class $\mathcal{G}$. Consequently, it is possible to compute the average fitness of $c$ against the population, that is given by:
\begin{align}
  \label{eq:FittnessChoiceMechGame}
  F(c) = \int P_{M}(c') \ F(c,c') \text{ d} c' = \int \int P_{\mathcal{G}}(G) \  P_{M}(c') \  F_G(c,c') \text{ d} c' \text{ d} G  \,,
\end{align}
where $P_{M}(c')$ is the probability of
encountering a co-player with choice mechanism $c'$.

Meta-games are then abstract models for the evolutionary competition between choice
mechanisms in interactive decision making contexts.
Standard notions of evolutionary game theory apply to meta-games as well. A choice
mechanism $c$ is a strict Nash equilibrium if $F(c,c) > F(c',c)$ for all $c'$; it is
evolutionarily stable if for all $c'$ either (i) $F(c,c) > F(c',c)$ or (ii) $F(c,c) = F(c',c)$
and $F(c,c') > F(c',c')$; it is neutrally stable if for all $c'$ either (i) $F(c,c) > F(c',c)$
or (ii) $F(c,c) = F(c',c)$ and $F(c,c') \ge F(c',c')$
\citep{Maynard-Smith1982:Evolution-and-t}. Similarly, evolutionary dynamics can be applied to
meta-games. We will later turn towards replicator dynamics
\citep{TaylorJonker1978:Evolutionary-St} and replicator mutator dynamics
\citep[e.g.][]{Nowak2006:Evolutionary-Dy}.
%\todo{footnote with caveat about agent-level processes}

\section{The Model} \label{sec:model}

At first, the main subjective utilities that are going to compete against each other for the survival of the fittest are essentially two:
\begin{enumerate}
\item the \textit{objective} utility, defined by: for all $G \in \mathcal{G}$,
$$\text{obj}^G(a,a')=\pi^G(a,a');$$
\item the \textit{regret}, defined by: for all $G \in \mathcal{G}$,
$$\text{reg}^G(a,a') =\pi^G(a,a')- \max_{a''\in A} \pi^G(a'',a').$$
\end{enumerate}

For a start, the subjective beliefs that we take into consideration are also two:
\begin{enumerate}
\item $prc$: a precise uniform belief $\overline{\mu}$ such that $\overline{\mu}(a)=\frac{1}{|A|}$ for all $a\in A$;
\item $imp$: a maximally imprecise belief $\overline{\Gamma}=\Delta(A)$.
\end{enumerate}

Although a thorough discussion of this issue goes beyond the scope of this work, let us say that these two kinds of belief underlie two different and alternative views on uncertainty. In front of uncertain events, a strict Bayesian will always form a precise belief, specified by a \textit{single} probability $\mu$. In the absence of any information about future uncertain events, the Bayesian would then invoke the \textit{principle of insufficient reason}, and accordingly choose a uniform probability over the possible outcomes. On the contrary, more recently many authors (REFERENCES) argued against the obligation of representing a belief by means of a single probability measure, opposite to the Bayesian prescriptions. They argue instead in favor of a more general account, according to which uncertainty can be unmeasurable, and represented by a (closed and convex) set of probabilities. This line of thought has been originated in decision theory by the famous Ellsberg's paradoxes (REF), and appears extremely relevant
in game-theoretical contexts too. Indeed, in a recent paper \citet{BattCerrMM15} write:

\noindent In evolutionary game theory, for instance, players will obviously face
uncertainty about the composition of the population that they are part of, and consequently
about the co-player that they are randomly paired with at each round and about the co-player's
action. In case of complete lack of information, a non-Bayesian player would thus entertain maximal unmeasurable uncertainty, i.e., a maximally imprecise belief. 
As already anticipated,  we will see that the way agents can form beliefs, and the possibility of holding imprecise beliefs in particular, may have a fundamental impact on their evolutionary success.
